Implicit Neural Representation for Potential Field Geophysics

### 2023-10-09 intro.tex
Write ML background

% indicates that an original and coherent argument is being developed and that the research is making a contribution to existing literature
% outline the topic, justify the research, identify research aims and outline thesis structure

@@

The subsurface continuation of exposed geology can be extrapolated and interpolated using structures mapped in surveys of the Earth's naturally occuring potential fields.
Potential field surveys offer low-cost investigations over a large spatial extent, and routinely used for geological mapping and interpretation \parencite{nabighian75thAnniversaryHistorical2005}
For this reason, magnetic field surveys are commonly used by geophysicists for geological mapping and interpretation toward understanding the Earth in three dimensions.
Scattered surveys are widely used when interpolated to grid rasters, and interpreted individually or integrated with many other datasets for predicting the subsurface for processes such as numerical modelling and inversion.
In these tasks, the spatial resolution of the grid is a critical factor in their ability to define geological detail \parencite{islesGeologicalInterpretationAeromagnetic2018}.
The role of high-resolution details in these geophysical methods is becoming more important to exploration as resources are increasingly sought from below sedimentary cover.

@@

The above aims are achieved using case studies using open access aeromagnetic data provided by Geoscience Australia.
Inference on real-world data presents many challenges \parencite{nikolenkoSyntheticDataDeep2021,tremblayTrainingDeepNetworks2018}.
Despite this, case studies are critical in promoting the adoption of applications in geophysics, by demonstrating a methods suitability to the data used within the task of interest.

\section{Aeromagnetic surveys}
\label{sec:introgeo}
% how are they gathered (surveys)
Airborne surveys of the Earth's magnetic or gravitational potential fields are routinely acquired at a range of scales at different stages of exploration, from pre-competitive regional data to stimulate exploration \parencite{howardAirborneGeophysicalCoverage2004}, to targeted investigations by the exploration industry at the prospect scale of up to several square kilometres.
Surveys may also be collected on the ground, or less commonly in exploration, by ship or satellite, with most processing methods and interpretation shared between all forms of acquisition.
Aeromagnetic surveys are among the lowest cost geophysical method \parencite{dentithGeophysicsMineralExploration2014}, however the cost rapidly increases when surveying at higher resolution due to overheads in flight line sampling.
When acquiring data, aircraft fly straight transect lines at a safe elevation above topography, known as the topography drape.
Surveys are flown with a heading in which data will be acquired along line.
For regional and contemporary surveys this is recommended to be North-South, however early practice was to survey perpendicular to the dominant geological strike, under the expectation of sampling the greatest frequency of change in magnetic intensity \parencite{islesGeologicalInterpretationAeromagnetic2018}.
Higher frequencies with these data diminish with increasing source-sensor distance, therefore the lower the height, the more high-frequency components will be captured by the sensor.
Acquisition along flight lines is sampled at a rate of \qty{10}{\hertz} or greater, which corresponds to an interval of \qty{7}{\m} or shorter at nominal flight speeds of \qty{70}{\m\per\s} \parencite{goodwinAirborneMagneticRadiometric2023}.
The spacing between lines is a factor of the scale of investigation and the cost of acquisition, and regional pre-competitive data in Australia targets state-wide coverage at \qty{400}{\m}, which is generally regarded as the upper limit of usefulness to exploration \parencite{howardAirborneGeophysicalCoverage2004}.
These low-resolution surveys, funded by government geoscience agencies, are a key driver in greenfields exploration and are released as open file data.
Higher-resolution surveys over prospective areas fly lines at closer line spacings, of the order \qty{100}{\m}, in order to resolve greater detail.
The importance of these parameters in relation to resolvable detail will be outlined shortly.
The collected data may contain millions of samples scattered in three dimensions, and interpretation or processing often relies on regularisation to a quantised grid.

\subsection{Survey data regularisation}
\label{sec:introgrids}
% - how are they transformed (gridding)
Survey data can be interpreted as one dimensional transects, but more commonly they are regularised to a two-dimensional grid raster for interpretation in a process termed gridding.
The resulting interpolated array contains uniformly spaced nodes and is referred to as a grid.
Grids are raster data, where each pixel stores a scalar measurement of a specific property.
In geophysics and throughout this thesis, these are referred to as cells, and the property is a measure of the strength of the magnetic field at a specific location.
The field strength measured is the Total Magnetic Intensity (TMI), which is the vector sum of each directional vector of the potential \parencite{blakelyPotentialTheoryGravity1996}.
This scalar value is recorded in nanoTesla and may be negative with an unbounded range that can exceed \qty{50,000}{\nano\tesla} in the Earth's environment.
The location recorded at a sample point is the geographic latitude and longitude, and altitude, acquired with high-precision differential GPS, alongside radar altimeter elevation.
These data may be transformed to a flat projected system, where spatial dimensions are expressed in meters, used throughout this thesis.
There are numerous conventional gridding methods, and novel methods are frequently proposed in the literature.
The simplest use surrounding samples for interpolation, such as choosing the closest sample value or applying a transformation function to the neighbouring samples.
Of those most commonly used in potential field geophysics, these are minimum curvature \parencite{briggsMachineContouringUsing1974}, splines \parencite{bhattacharyyaBicubicSplineInterpolation1969,shureHarmonicSplinesGeomagnetic1982,smithGriddingContinuousCurvature1990}, and equivalent sources \parencite{dampneyEquivalentSourceTechnique1969, solerBetterStrategyInterpolating2020}.
Splines are used in bi-directional gridding, which is well-suited for aeromagnetic surveys, where the sample data are first interpolated in the line parallel direction, and the intermediate product is interpolated in the line perpendicular direction \parencite{dentithGeophysicsMineralExploration2014}.
A method known as kriging is used in other areas of geoscience, such as geochemical mapping, and is an interpolator which makes use of statistical properties of the known distribution \parencite{hansenInterpretiveGriddingAnisotropic1993,davis1986statistics}.
Some recently proposed gridding methods include the methods of \textcite{naprstekNewMethodInterpolating2019}, \textcite{xuGravityAnomalyReconstruction2019}, or \textcite{chenPotentialFieldData2022}.
Each interpolation method attempts to predict regularly spaced grids by leveraging properties intrinsic to the data, with the possible explicit integration of \emph{a priori} knowledge such as geophysical laws or geological context.
A review of spatial interpolation methods is provided in \textcite{liReviewComparativeStudies2011}.

\subsection{A definition for geophysical resolution}
In all interpolated grids, the spatial dimension of each cell is referred to as the cell size, with the unit of metre used in this thesis.
As previously stated, resolvable detail in geophysical surveys is a product of the distance between the causative magnetic body and sensor (height), and the line spacing \parencite{islesRelationshipsGeologicalResolution1992,islesGeologicalInterpretationAeromagnetic2018,dentithGeophysicsMineralExploration2014}.
Height is easily controlled by flying the lowest safe and sufficiently smooth topography drape, and along-line sampling is performed at a constant high rate, so the key factor controlling resolution in aeromagnetic survey design is the line spacing.
Further survey design considerations for airborne geophysics are outlined in \textcite{goodwinAirborneMagneticRadiometric2023,islesGeologicalInterpretationAeromagnetic2018,reidAeromagneticSurveyDesign1980}.
Each gridding method includes a final cell size parameter selectable by the geophysicist, with a long-standing guideline of one third to one fifth the line spacing.
This sufficiently preserves detail in areas of high sampling, while avoiding excessive interpolated cells in areas of low sampling, which may lead to artefacts.
Because of the sampling frequency of nominally \qty{7}{\m} in the flight line direction, and  up to \qty{400}{\m} in the line perpendicular direction, the resulting grid resolution will be anisotropic at best, and uniformly low-resolution at worst.
An important property of sampled data is the Nyquist frequency, \[f_{Nyquist} = f_{sampling} / 2.\]
This is the lowest frequency at which a bandwidth limited signal may be sufficiently sampled, beyond which higher frequencies will be incorrectly recorded as aliasing.
An aliased signal contains spurious low-frequency features caused by the sampling, not the underlying signal.
This limiting frequency applies to each sampling direction in surveys, as well as the cell size of the grid.
The smallest resolvable feature in each of these can be spatially described by wavelength, as the inverse of frequency.
In low-resolution aeromagnetic surveys of \qty{400}{\m} line spacing, strong variations of magnetic intensity in the line-perpendicular direction  that occur over a distance shorter than \qty{800}{\m} will be aliased.
When gridded at \qty{80}{\m} cell size, the corresponding limit on resolvable detail is features as small as \qty{160}{\m}.
% This choice of cell size will effectively filter out features in the line-parallel direction that vary strongly, from those possibly present at \qty{14}{\m} through to \qty{160}{\m}.
Depending on the complexity of the geological domain, this may cause spatial aliasing in one direction.
This is visible in grids where features take on a blocky or stepped appearance, referred to as a beading artefact, or generally in this thesis as a low-resolution artefact.
Low-resolution gridded aeromagnetic data refers to surveys with a wide line spacing causing a lack of high-frequency details in the line-perpendicular direction, resulting in potential artefacts, and overly smooth or missing details in grid data.

@@

This can be increased by:
\begin{enumerate}
    \item{} High frequencies, caused by geology, being present in the potential field,
    \item{} Capturing these high frequencies during surveying, by sampling at close line spacing and small source-sensor separation,
    \item{} Interpolating sample data with a small enough spatial cell size to preserve these frequencies.
\end{enumerate}

% - How are they used (geo interp, modelling, filtering) (detailed)
% levelling
% geological interpretation
% modellin
%

@@

Geoscience data faces a number of challenges distinct from natural images. These have been outlined in \textcite{karpatneMachineLearningGeosciences2019}, and relevant to this work are:

\begin{itemize}
    \item{} Spatio-Temporal Structure: Aeromagnetic grids are spatially corellated, neighbouring cells contain similar values arising from the same causative volume.
    \item{} Multi-Source Multi-Resolution Data: One location may be covered by mutiple aeromagnetic surveys with different resolutions, or the same survey may be presented with different levels of processing.
    \item{} Poor Quality of Data: Sample data are subject to numerous sources of noise, which combine to become a significant fraction of the gridded data value, or be present as aliased frequencies.
    \item{} Small Sample Size: While total aeromagnetic coverage in Australia is high, the high-resolution extents are spatially limited and biased by geological domain.
\end{itemize}

One approach to overcoming the challenges faced in real-world geoscience data is to use synthetic data.
During this thesis, a large collection of synthetic data became available, containing realistic 3D geological voxel models and 2D geophysical forward models \parencite{jessellNoddyverseMassiveData2022}.
These data present a large amount of uniformly sampled and noise free grids from an unbiased set of geological domains.

@@

The rate at which the power of the component frequencies of the potential field diminish with increasing distance is well known.
Back-calculation can be made to predict the power of these frequencies at a smaller separation, and this is termed downward continuation \textcite{bullardDeterminationMassesNecessary1948,blakelyPotentialTheoryGravity1996}.
The inverse using a larger distance is termed upward continuation, and is useful for reducing potentially unwanted near-surface magnetic detail, or levelling data to a common datum \parencite[e.g.\ open file magnetic data in Australia is available levelled to AWAGS][]{mintyAirborneGeophysicalMapping2011}.
Downward continuation offers the benefit of high-resolution data from low-resolution sampling, and is therefore highly studied \parencite{zuoDownwardContinuationTransformation2020,liStableDownwardContinuation2023,fediStableDownwardContinuation2002,yeHighprecisionDownwardContinuation2022,zhangNumericalSolutionsMeanValue2018a,guoPotentialFieldContinuation2020,gangImprovedStableDownward2018,pilkingtonPotentialFieldContinuation2017}.
However, due to the inclusion of sensor noise, the practicable limit of numerical methods is six times the low-resolution cell size.
Recent approaches in machine learning that overcome this limit will be discussed in \Cref{sec:introml}, alongside alternative methods for enhancing aeromagnetic survey resolution.

\section{Machine learning for enhancing resolution}
% Representation learning...

\label{sec:introml}
Deep learning is a field of contemporary machine learning, where multiple stages of features learning are undertaken between the input and output.
The current generation of deep learning is reported to have begun in 2006 \parencite{dengDeepLearningMethods2014,Goodfellow-et-al-2016}, and has pervaded many industries and research fields.
The network architectures encountered in this thesis include convolutional neural networks (CNN), multilayer perceptrons (MLP), and generative adversarial networks (GAN).

CNN methods have long been well suited for computer vision tasks such as image classification and, more recently, super-resolution.
In an image raster, neighbouring pixels define simple features, and these features combine in the wider image to define complex features.
CNNs excel in deep learning tasks for these data, due in part to the structure of the receptive field of the network.
The receptive field is formed from stacked layers of small convolutional kernels, typically with a size of \numproduct{3 x 3} cells for super-resolution CNNs.
These low-level kernels learn simple functions, which are convolved with deeper layers to learn increasingly complex patterns across a larger image extent, while widening the receptive field.
A review of deep learning CNN is provided in \textcite{goodfellowDeepLearning2016}.
% Once trained, the learnt functions are conditioned to transform low-resolution features into high-resolution features, using the neighbouring input values of the low-resolution grid in the context of filters learnt from the training data.

@@

% \begin{itemize}
%     \item GAN    
%           %  are an architecture making use of two competing neural networks.
%           % Highly suited for generative modelling, 
%     \item Introduce different types of networks and their applications in geoscience and adjacent areas
%           There has been active area of computer vision research in developing various deep learning network architectures (*refs) to enhance the resolution of images (super-resolution) based on previously seen patterns.
%           These are described in \Cref{sec:introgeoml}
%     \item * talk about different superresolution methods that have been effective for some applications with refs*, and introduce learning implicit representation from imagery and their application for improving resolution *** I think it's useful to link this resolution improvement in geophysics as imputation problem.
%           deep learning methods that consider the characteristics of geophysics potential fields data.
%     \item revisit or move Deep learning downward continuation
%     \item *needs a bit of literature review on what's been done before in this space in geoscience***
%     \item to the best of my knowledge, this use of super-resolution ***
% \end{itemize}

@@

>>>>

@@

The subsurface continuation of exposed geology can be extrapolated and interpolated using structures mapped in surveys of the Earth's naturally occuring potential fields.
Potential field surveys offer low-cost investigations over a large spatial extent, and routinely used for geological mapping and interpretation \parencite{nabighian75thAnniversaryHistorical2005}
For this reason, magnetic field surveys are commonly used by geophysicists for geological mapping and interpretation toward understanding the Earth in three dimensions.
Scattered surveys are widely used when interpolated to grid rasters, and interpreted individually or integrated with many other datasets for predicting the subsurface for processes such as numerical modelling and inversion.
In these tasks, the spatial resolution of the grid is a critical factor in their ability to define geological detail \parencite{islesGeologicalInterpretationAeromagnetic2018}.
The role of high-resolution details in these geophysical methods is becoming more important to exploration as resources are increasingly sought from below sedimentary cover.
Resolution enhancement has seen recent success from the use of deep learning \parencite{moserHitchhikerGuideSuperResolution2023}, however it is not understood if these methods could extend to geophysics data, which are highly distinct from natural images.

@@

The above aims are achieved using case studies with open access aeromagnetic data provided by Geoscience Australia.

\section{Aeromagnetic surveys}
\label{sec:introgeo}
Airborne surveys of the Earth's magnetic or gravitational potential fields are routinely acquired at a range of scales at different stages of exploration, from pre-competitive regional data to stimulate exploration \parencite{howardAirborneGeophysicalCoverage2004}, to targeted investigations by the exploration industry at the prospect scale of up to several square kilometres.
Surveys may also be collected on the ground, or less commonly in exploration, by ship or satellite, with most processing methods and interpretation shared between all forms of acquisition.
Aeromagnetic surveys are among the lowest cost geophysical method \parencite{dentithGeophysicsMineralExploration2014}, however the cost rapidly increases when surveying at higher resolution due to overheads in flight line sampling.
When acquiring data, aircraft fly straight transect lines at a safe elevation above topography, known as the topography drape.
Surveys are flown with a heading in which data will be acquired along line.
For regional and contemporary surveys this is recommended to be North-South, however early practice was to survey perpendicular to the dominant geological strike, under the expectation of sampling the greatest frequency of change in magnetic intensity \parencite{islesGeologicalInterpretationAeromagnetic2018}.
Higher frequencies with these data diminish with increasing source-sensor distance, therefore the lower the height, the more high-frequency components will be captured by the sensor.
Acquisition along flight lines is sampled at a rate of \qty{10}{\hertz} or greater, which corresponds to an interval of \qty{7}{\m} or shorter at nominal flight speeds of \qty{70}{\m\per\s} \parencite{goodwinAirborneMagneticRadiometric2023}.
The spacing between lines is a factor of the scale of investigation and the cost of acquisition, and regional pre-competitive data in Australia targets state-wide coverage at \qty{400}{\m}, which is generally regarded as the upper limit of usefulness to exploration \parencite{howardAirborneGeophysicalCoverage2004}.
These low-resolution surveys, funded by government geoscience agencies, are a key driver in greenfields exploration and are released as open file data.
Higher-resolution surveys over prospective areas fly lines at closer line spacings, of the order \qty{100}{\m}, in order to resolve greater detail.
The importance of these parameters in relation to resolvable detail will be outlined shortly.
The collected data may contain millions of samples scattered in three dimensions, and interpretation or processing often relies on regularisation to a quantised grid.

\subsection{Survey data regularisation}
\label{sec:introgrids}
Survey data can be interpreted as one dimensional transects, but more commonly they are regularised to a two-dimensional grid raster for interpretation in a process termed gridding.
The resulting interpolated array contains uniformly spaced nodes and is referred to as a grid.
Grids are raster data, where each pixel stores a scalar measurement of a specific property.
In geophysics and throughout this thesis, these are referred to as cells, and the property is a measure of the strength of the magnetic field at a specific location.
The field strength measured is the Total Magnetic Intensity (TMI), which is the vector sum of each directional vector of the potential \parencite{blakelyPotentialTheoryGravity1996}.
This scalar value is recorded in nanotesla and may be negative with an unbounded range that can exceed \qty{50,000}{\nano\tesla} in the Earth's environment.
The location recorded at a sample point is the geographic latitude and longitude, and altitude, acquired with high-precision differential GPS, alongside radar altimeter elevation.
These data may be transformed to a flat projected system, where spatial dimensions are expressed in meters, used throughout this thesis.
There are numerous conventional gridding methods, and novel methods are frequently proposed in the literature.
The simplest use surrounding samples for interpolation, such as choosing the closest sample value or applying a transformation function to the neighbouring samples.
Of those most commonly used in potential field geophysics, these are minimum curvature \parencite{briggsMachineContouringUsing1974}, splines \parencite{bhattacharyyaBicubicSplineInterpolation1969,shureHarmonicSplinesGeomagnetic1982,smithGriddingContinuousCurvature1990}, and equivalent sources \parencite{dampneyEquivalentSourceTechnique1969, solerBetterStrategyInterpolating2020}.
Splines are used in bi-directional gridding, which is well-suited for aeromagnetic surveys, where the sample data are first interpolated in the line parallel direction, and the intermediate product is interpolated in the line perpendicular direction \parencite{dentithGeophysicsMineralExploration2014}.
A method known as kriging is used in other areas of geoscience, such as geochemical mapping, and makes use of statistical properties of the known distribution \parencite{hansenInterpretiveGriddingAnisotropic1993,davis1986statistics}.
Some recently proposed gridding methods include the methods of \textcite{naprstekNewMethodInterpolating2019}, \textcite{xuGravityAnomalyReconstruction2019}, or \textcite{chenPotentialFieldData2022}.
Each interpolation method attempts to predict regularly spaced grids by leveraging properties intrinsic to the data, with the possible explicit integration of \emph{a priori} knowledge such as geophysical laws or geological context.
A review of spatial interpolation methods is provided in \textcite{liReviewComparativeStudies2011}.

\subsection{A definition for geophysical resolution}
In interpolated grids the spatial dimension of each cell is referred to as the cell size, with a unit of metre used in this thesis.
As previously stated, resolvable detail in geophysical surveys is a product of the distance between the causative magnetic body and sensor (height), and the line spacing \parencite{islesRelationshipsGeologicalResolution1992,islesGeologicalInterpretationAeromagnetic2018,dentithGeophysicsMineralExploration2014}.
Height is easily controlled by flying the lowest safe and sufficiently smooth topography drape, and along-line sampling is performed at a constant high rate, so the key factor controlling resolution in aeromagnetic survey design is the line spacing.
Further survey design considerations for airborne geophysics are outlined in \textcite{goodwinAirborneMagneticRadiometric2023,islesGeologicalInterpretationAeromagnetic2018,reidAeromagneticSurveyDesign1980}.
Each gridding method includes a final cell size parameter selectable by the geophysicist, with a long-standing guideline of one third to one fifth the line spacing.
This sufficiently preserves detail in areas of high sampling, while avoiding excessive interpolated cells in areas of low sampling, which may lead to artefacts.
Because of the sampling frequency of nominally \qty{7}{\m} in the flight line direction, and  up to \qty{400}{\m} in the line perpendicular direction, the resulting grid resolution will be anisotropic at best, and uniformly low-resolution at worst.
An important property of sampled data is the Nyquist frequency, \[f_{Nyquist} = f_{sampling} / 2.\]
This is the lowest frequency at which a bandwidth limited signal may be sufficiently sampled, beyond which higher frequencies will be incorrectly recorded as aliasing.
An aliased signal contains spurious low-frequency features caused by the sampling, not the underlying signal.
This limiting frequency applies to each sampling direction in surveys, as well as the cell size of the grid.
The smallest resolvable feature in each of these can be spatially described by wavelength, as the inverse of frequency.
In low-resolution aeromagnetic surveys of \qty{400}{\m} line spacing, strong variations of magnetic intensity in the line-perpendicular direction  that occur over a distance shorter than \qty{800}{\m} will be aliased.
When gridded at \qty{80}{\m} cell size, the corresponding limit on resolvable detail is features as small as \qty{160}{\m}.
% This choice of cell size will effectively filter out features in the line-parallel direction that vary strongly, from those possibly present at \qty{14}{\m} through to \qty{160}{\m}.
Depending on the complexity of the geological domain, this may cause spatial aliasing in one direction.
This is visible in grids where features take on a blocky or stepped appearance, referred to as a beading artefact, or generally in this thesis as a low-resolution artefact.
Low-resolution gridded aeromagnetic data refers to surveys with a wide line spacing causing a lack of high-frequency details in the line-perpendicular direction, resulting in potential artefacts, and overly smooth or missing details in grid data.

@@

Resolution can be increased by:
\begin{enumerate}
    \item{} High frequencies, caused by geology, being present in the potential field,
    \item{} Capturing these high frequencies during surveying, by sampling at close line spacing and small source-sensor separation,
    \item{} Interpolating sample data with a small enough spatial cell size to preserve these frequencies.
\end{enumerate}

% - How are they used (geo interp, modelling, filtering) (detailed)
% levelling
% geological interpretation
% modelling
%

@@

Geoscience data faces a number of challenges distinct from natural images. These have been outlined in \textcite{karpatneMachineLearningGeosciences2019}, and relevant to this work are:
\begin{itemize}
    \item{} Spatio-Temporal Structure: Aeromagnetic grids are spatially corellated, neighbouring cells contain similar values arising from the same causative volume.
    \item{} Multi-Source Multi-Resolution Data: One location may be covered by mutiple aeromagnetic surveys with different resolutions, or the same survey may be presented with different levels of processing.
    \item{} Poor Quality of Data: Sample data are subject to numerous sources of noise, which combine to become a significant fraction of the gridded data value, or be present as aliased frequencies.
    \item{} Small Sample Size: While total aeromagnetic coverage in Australia is high, the high-resolution extents are spatially limited and biased by geological domain.
\end{itemize}

Inference on real-world data presents many challenges \parencite{nikolenkoSyntheticDataDeep2021,tremblayTrainingDeepNetworks2018}.
Despite this, case studies are critical in promoting the adoption of applications in geophysics, by demonstrating a methods suitability to the data used within the task of interest.
As such, case studies are investigated for each aim within this thesis.

@@

The rate at which the power of the component frequencies of the potential field diminish with increasing distance is well known.
Back-calculation can be made to predict the power of these frequencies at a smaller separation, and this is termed downward continuation \textcite{bullardDeterminationMassesNecessary1948,blakelyPotentialTheoryGravity1996}.
The inverse using a larger distance is termed upward continuation, and is useful for reducing potentially unwanted near-surface magnetic detail, or levelling data to a common datum \parencite[e.g.\ open file magnetic data in Australia is available levelled to the AWAGS datum][]{mintyAirborneGeophysicalMapping2011}.
Downward continuation offers the benefit of high-resolution data from low-resolution sampling, and is therefore highly studied \parencite{zuoDownwardContinuationTransformation2020,liStableDownwardContinuation2023,fediStableDownwardContinuation2002,yeHighprecisionDownwardContinuation2022,zhangNumericalSolutionsMeanValue2018,guoPotentialFieldContinuation2020,gangImprovedStableDownward2018,pilkingtonPotentialFieldContinuation2017}.
However, due to the inclusion of sensor noise, the practicable limit of numerical methods is six times the low-resolution cell size.
Recent approaches in machine learning have overcome this limit, alongside alternative methods for enhancing aeromagnetic survey resolution.

\subsection{Deep learning context}
\label{sec:introml}
Machine learning (ML) is a very broad family of techniques for learning functions from data.
Artificial neural networks are a subset of techniques in ML, inspired by a model of the biological neuron \parencite{bishopNeuralNetworksPattern1995}.
It has been established that if sufficient conditions are met, neural networks can approximate any function \parencite{hornikMultilayerFeedforwardNetworks1989}.
A mathematical model of a neuron forms the basis of these networks, and the number of neurons within the network and their arrangement are one primary descriptor of a neural network. 
Neurons are paremeterised by a multiplicative weight term, an added bias term, and an activation function to scale the result.
Each neuron applies these operations to the input, which may either be the input data, or the output of earlier neurons.
The objective of training a neural network is to set the values for these for each neuron in the network, such that the trained network (or model) generalises to novel data.
This is performed by back propogation \parencite{rumelhartLearningRepresentationsBackpropagating1988}, a technique to navigate toward the lowest error solution by calculating the rate of improvement resulting from parameter updates.
A practical introduction to the practice of training machine learning is available from \textcite{stevensDeepLearningPyTorch2020}, while a more formal introduction is provided in \textcite{bishopPatternRecognitionMachine2006}.
Many activation functions have been proposed, with the rectified linear unit (ReLU) being used in most tasks throughout this thesis.
The activation function performs scaling but in many classes of activation function it also performs the important task of scaling the result to zero, which can be interpreted as the lack of a specific feature being present \parencite{williamsLogicActivationFunctions1986}.
However, recent work on periodic activation functions such as the sinusoid \parencite{sitzmann2019siren} enable the networks used in Cref{ch:paper3}.
The other primary descriptor of a neural network is its architecture, refering to the configuration of its intermediate layers between the input and output.
These architectures are described below.

If the network architecture provides the capacity to store and retrieve learning, the data and training architecture provide the actual instruction and delivery.
There are broadly two classes of training; supervised and unsupervised learning. 
The super-resolution and representation learning tasks in this thesis are both supervised tasks, where a reference is provided for the network to calculate an objective function of its performance on.
Unsupervised learning does not have a ground truth or target, and instead learns to identify patterns within the data for tasks in clustering, density estimation, and visualisation \parencite{bishopPatternRecognitionMachine2006}.

Deep learning is a field of contemporary machine learning, where multiple layers of feature learning are undertaken between the input to a method and the output.
The current generation of deep learning is reported in review articles to have begun in 2006 \parencite{dengDeepLearningMethods2014,goodfellowDeepLearning2016a}, and has pervaded many industries and research fields.
These review articles contain thorough background on the implementation and practice of various architectures for undertaking deep learing, as well as earlier machine learning or numerical models that can address tasks now approached with deep learning.
The deep learning tasks explored in this thesis are super-resolution (SR) for the objective of upsampling resolution enhancement, and implicit neural representation (INR).
The network architectures implemented in this thesis to address the tasks include convolutional neural networks (CNN), generative adversarial networks (GAN), and multilayer perceptrons (MLP).
A brief introduction to these topics is provided here.

In an image raster, neighbouring pixels define simple features, and these features combine in the wider image to define complex features.
A CNN is formed from stacked layers of small convolutional kernels, commonly in SR with the kernel size of \numproduct{3 x 3} cells.
While training, these low-level kernels learn simple functions, which are convolved with deeper layers to learn increasingly complex patterns across a larger image extent, while widening the receptive field.
CNN methods have long been well suited for raster data in computer vision tasks such as image classification \parencite[e.g.][]{simonyanVeryDeepConvolutional2015} and, more recently, resolution enhancement \parencite[e.g.][]{zhangResidualDenseNetwork2018}.
The earliest application of CNN to SR was demonstrated by \textcite{dongLearningDeepConvolutional2014}, where convolutional layers throughout the depth of a network are used for low-resolution patch extraction, feature mapping, and reconstruction of super-resolved images.
The outputs of each convolutional layer are similar to hand-crafted filters or the feature dictionaries in earlier example-based super-resolution methods \parencite{freemanExamplebasedSuperresolution2002}, but the learnt filters are more numerous and fully learned from the data with the deep learning approach.
Following the success of \parencite{dongLearningDeepConvolutional2014}, many iterative improvements have been made to CNN based SR\@.
These include SRCNN \parencite{dongImageSuperresolutionUsing2016}, RDN \parencite{zhangResidualDenseNetwork2018}, ESRGAN \parencite{wangESRGANEnhancedSuperresolution2018}, and others \parencite{ledigPhotorealisticSingleImage2017,limEnhancedDeepResidual2017}.
% While each of these works contribute unique features to CNN-based super-resolution, they follow a common stacked layer design.
% This comprises a set of initial input feature extraction convolutions, a sequence of convolutional blocks with activations (neurons), some number and arrangement of intermediate residual learning pathways (skip connections), and an upsampling block for feature upscaling and convolving latent features back to image space.
% Once trained, the learnt functions are conditioned to transform low-resolution features into high-resolution features, using the neighbouring input values of the low-resolution grid in the context of filters learnt from the training data.

@@

Generative Adversarial Networks \parencite{goodfellowGenerativeAdversarialNets2014} are a learning framework utilising two competing neural networks.
These are termed the generator and the discriminator, and can be any arbitrary networks linked by objective functions.
When best overall performance arises when each network is equally capable in its respective task, i.e.\ when the objective functions of each network are linked by the Nash equilibrium \parencite{salimansImprovedTechniquesTraining2016,lucicAreGANsCreated2018}.
Because the use of a discriminator allows an independent element for predictive capacity, GANs are highly suited for generative modelling, which includes the task of super-resolution.
However, it was considerably challenging to achieve a stable Nash equilibrium \parencite{salimansImprovedTechniquesTraining2016}, causing difficulty in training these networks, and to produce a diverse range of outputs.
Noise injection was later used in this framework as a method to drive diversity \parencite{karrasStylebasedGeneratorArchitecture2018,rakotonirinaESRGANFurtherImproving2020}.

%     \item * talk about different superresolution methods that have been effective for some applications with refs*, and introduce learning implicit representation from imagery and their application for improving resolution *** I think it's useful to link this resolution improvement in geophysics as imputation problem.

\section{Machine learning in geoscience}
\label{sec:introgeoml}
%     \item *needs a bit of literature review on what's been done before in this space in geoscience***
%     \item to the best of my knowledge, this use of super-resolution ***
% \end{itemize}

@@

\printbibliography{}
